---
title: "Analyzing Crime Patterns in Chicago"
format: pdf
editor: visual
---

```{r, echo=FALSE}
df <- read.csv('/Users/shreyapatchala/Desktop/Applied LM/Assignments/Final /Final/Chicago.crime.csv', row.names = 1)
```

## Introduction

The City of Chicago maintains a comprehensive database of criminal reports, including narcotics-related crimes such as marijuana possession (CANNABIS) and other offenses (NOTCANNABIS). This dataset, combined with demographic information from the 2010 US Census and the 2011 American Community Survey, offers a unique opportunity to analyze crime patterns across different neighborhoods. This report aims to explore key questions regarding the distribution of crime rates, the relationship between demographic factors and crime, and temporal trends.

## Exploratory Data Analysis

**Data Summary:\
**The dataset containscensus block groups in Chicago, with variables that include demographic information and crime data. It provides insights into how these factors vary across different neighborhoods and their potential relationship to crime rates. The crime data, spread over several years, offers an opportunity to explore patterns in crime, while demographic variables allow for the investigation of socio-economic and geographic factors that might influence crime trends.

-   **Population Data**: `poptotal`, `popwhite`, `popblack`, `popasian`

-   **Income Data**: `income.male`, `income.female`

-   **Age Data**: `age.male`, `age.female`

-   **Geographic Data**: `longitude`, `latitude`

-   **Crime Data**: Crime counts for two types—`CANNABIS` and `NOTCANNABIS`—are recorded for 2010, 2011 and 2012.

**Data Exploration:\
**NA values were removed from the dataset as they were sparse in number and did not contribute significantly to the overall analysis. Additionally, removing these NA values also eliminated rows where the population count was zero, which would have been irrelevant for future analysis. This cleaning step ensured that only meaningful data, with valid population values, was retained for further analysis.

```{r, echo=FALSE}
# Remove rows with any NA values
df<- na.omit(df)
```

```{r, echo=FALSE, fig.width=6, fig.height=4, fig.cap="Distribution of Predictors"}
# Load required libraries
library(ggplot2)
library(gridExtra)

# Plot 1: Histogram for poptotal
p1 <- ggplot(df, aes(x = poptotal)) +
  geom_histogram(binwidth = 500, fill = "purple", color = "black", alpha = 0.7) +
  labs(title = "Total Population", x = "Total Population", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 2: Histogram for popwhite
p2 <- ggplot(df, aes(x = popwhite)) +
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "White Population", x = "White Population", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 3: Histogram for popblack
p3 <- ggplot(df, aes(x = popblack)) +
  geom_histogram(binwidth = 200, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Black Population", x = "Black Population", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 4: Histogram for popasian
p4 <- ggplot(df, aes(x = popasian)) +
  geom_histogram(binwidth = 50, fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Asian Population", x = "Asian Population", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 5: Histogram for income.male
p5 <- ggplot(df, aes(x = income.male)) +
  geom_histogram(binwidth = 5000, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Male Income", x = "Male Income", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 6: Histogram for income.female
p6 <- ggplot(df, aes(x = income.female)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Female Income", x = "Female Income", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 7: Histogram for age.male
p7 <- ggplot(df, aes(x = age.male)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Male Age", x = "Male Age", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 8: Histogram for age.female
p8 <- ggplot(df, aes(x = age.female)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black", alpha = 0.7) +
  labs(title = "Female Age", x = "Female Age", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1))

# Combine all plots into one window
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 3)



```

```{r, echo=FALSE}
# Sum the total crimes (both kinds) for each block group
df$total_cannabis <- rowSums(df[, grep("CANNABIS", colnames(df))], na.rm = TRUE)
df$total_notcannabis <- rowSums(df[, grep("NOTCANNABIS", colnames(df))], na.rm = TRUE)

# Calculate the crime rate for cannabis and non-cannabis crimes
df$cannabis_crime_rate <- (df$total_cannabis / df$poptotal) * 1000
df$notcannabis_crime_rate <- (df$total_notcannabis / df$poptotal) * 1000

```

```{r, echo=FALSE}
# Calculate the mean crime rates for cannabis and non-cannabis crimes
mean_cannabis_rate <- mean(df$cannabis_crime_rate, na.rm = TRUE)
mean_notcannabis_rate <- mean(df$notcannabis_crime_rate, na.rm = TRUE)

# Identify block groups with higher or lower crime rates than the mean
df$high_cannabis <- ifelse(df$cannabis_crime_rate > mean_cannabis_rate, "High", "Low")
df$high_notcannabis <- ifelse(df$notcannabis_crime_rate > mean_notcannabis_rate, "High", "Low")

```

```{r, echo=FALSE}
# # rows with zero population
# zero_population <- df[df$poptotal == 0, ]
# print(zero_population)
```

```{r, echo=FALSE}
# Filter out rows where population is zero
df_clean <- df

# Now calculate crime rates on the cleaned data
df_clean$total_cannabis <- rowSums(df_clean[, grep("CANNABIS", colnames(df_clean))], na.rm = TRUE)
df_clean$total_notcannabis <- rowSums(df_clean[, grep("NOTCANNABIS", colnames(df_clean))], na.rm = TRUE)

df_clean$cannabis_crime_rate <- (df_clean$total_cannabis / df_clean$poptotal) * 1000
df_clean$notcannabis_crime_rate <- (df_clean$total_notcannabis / df_clean$poptotal) * 1000

# Calculate mean crime rates again
mean_cannabis_rate <- mean(df_clean$cannabis_crime_rate, na.rm = TRUE)
mean_notcannabis_rate <- mean(df_clean$notcannabis_crime_rate, na.rm = TRUE)

# Identify high/low crime rates
df_clean$high_cannabis <- ifelse(df_clean$cannabis_crime_rate > mean_cannabis_rate, "High", "Low")
df_clean$high_notcannabis <- ifelse(df_clean$notcannabis_crime_rate > mean_notcannabis_rate, "High", "Low")

```

**Geographic Distribution of Cannabis and Non-Cannabis Crime Rates in Chicago: Comparison to Mean Crime Rates:**\
In this analysis, we first derived the cannabis and non-cannabis crime rates per 100 population for each block group. We then classified each block group as either having "high" or "low" crime rates based on whether their crime rates were above or below the mean crime rate for each respective category. The comparison to the mean helped identify areas with significantly higher or lower crime rates for both cannabis-related and non-cannabis-related offenses.

```{r, echo=FALSE, fig.width=8, fig.height=7, fig.cap="Geographic Distribution of Crime Rates in Chicago", out.width="100%" }
# PART A)
# Load necessary libraries
library(ggplot2)
library(gridExtra)

# Plot 1: Geographic distribution of block groups with high or low cannabis crime rates
p1 <- ggplot(df_clean, aes(x = longitude, y = latitude, color = high_cannabis)) +
  geom_point() +
  labs(title = "Geographic Distribution of Cannabis Crime Rates",
       x = "Longitude", y = "Latitude") +
  scale_color_manual(values = c("Low" = "blue", "High" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 11))  # Reduce title size

# Plot 2: Geographic distribution of block groups with high or low non-cannabis crime rates
p2 <- ggplot(df_clean, aes(x = longitude, y = latitude, color = high_notcannabis)) +
  geom_point() +
  labs(title = "Geographic Distribution of Non-Cannabis Crime Rates",
       x = "Longitude", y = "Latitude") +
  scale_color_manual(values = c("Low" = "blue", "High" = "green")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 11))  # Reduce title size

# Combine both plots vertically
grid.arrange(p1, p2, nrow = 2)


```

Given that both cannabis-related and non-cannabis-related crime rates are evenly distributed across the city, as observed through their geographic spread along longitude and latitude, there appears to be no clear geographic concentration or clusters of block groups with significantly higher or lower crime rates. This suggests that crime patterns for both types of offenses are relatively uniform across the city, rather than being concentrated in specific neighborhoods.

**Crime Rate Variations and the Relationship Between Cannabis and Non-Cannabis Crimes Across Wards and Community Areas**

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# PART B)
# Load necessary packages
library(dplyr)

# Group by Ward and calculate average cannabis and non-cannabis crime rates
crime_rate_by_ward <- df_clean %>%
  group_by(Ward) %>%
  summarize(
    avg_cannabis_crime_rate = mean(cannabis_crime_rate, na.rm = TRUE),
    avg_notcannabis_crime_rate = mean(notcannabis_crime_rate, na.rm = TRUE)
  )

# Group by Community Area and calculate average cannabis and non-cannabis crime rates
crime_rate_by_community_area <- df_clean %>%
  group_by(Community.Area) %>%
  summarize(
    avg_cannabis_crime_rate = mean(cannabis_crime_rate, na.rm = TRUE),
    avg_notcannabis_crime_rate = mean(notcannabis_crime_rate, na.rm = TRUE)
  )
```

```{r, echo=FALSE, fig.width=8, fig.height=5, fig.cap="Crime Rate Variations by Ward and Community Area", out.width="100%" }

# Install and load necessary libraries
library(ggplot2)
library(gridExtra)

# Plot 1: Average Cannabis Crime Rate by Ward
p1 <- ggplot(crime_rate_by_ward, aes(x = factor(Ward), y = avg_cannabis_crime_rate)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Avg Cannabis Crime Rate by Ward", x = "Ward", y = "Avg Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5), plot.title = element_text(size = 8))  # Adjusted angle and size

# Plot 2: Average Non-Cannabis Crime Rate by Ward
p2 <- ggplot(crime_rate_by_ward, aes(x = factor(Ward), y = avg_notcannabis_crime_rate)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Avg Non-Cannabis Crime Rate by Ward", x = "Ward", y = "Avg Non-Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5), plot.title = element_text(size = 8))  # Adjusted angle and size

# Plot 3: Average Cannabis Crime Rate by Community Area
p3 <- ggplot(crime_rate_by_community_area, aes(x = factor(Community.Area), y = avg_cannabis_crime_rate)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Avg Cannabis Crime Rate by Community Area", x = "Community Area", y = "Avg Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 60, hjust = 0.5, size = 5), plot.title = element_text(size = 8))  # Adjusted angle and size

# Plot 4: Average Non-Cannabis Crime Rate by Community Area
p4 <- ggplot(crime_rate_by_community_area, aes(x = factor(Community.Area), y = avg_notcannabis_crime_rate)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Avg Non-Cannabis Crime Rate by Community Area", x = "Community Area", y = "Avg Non-Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 60, hjust = 0.5, size = 5), plot.title = element_text(size = 8))  # Adjusted angle and size

# Arrange all plots in the same window using grid.arrange()
grid.arrange(p1, p2, p3, p4, ncol = 2)

```

The crime rate patterns across Wards and Community Areas appear quite similar. Both levels show comparable trends in cannabis and non-cannabis crime rates, with some geographic areas exhibiting consistently higher crime rates.

Relationship Between Cannabis and Non-Cannabis Crimes at these levels:

\vspace{-5mm}}$$
\text{Avg Cannabis Crime Rate}_\text{Ward} = \beta_0 + \beta_1 \cdot \text{Avg Non-Cannabis Crime Rate}_\text{Ward} + \epsilon
$$ \vspace{-4mm} $$
\text{Avg Cannabis Crime Rate}_\text{Community Area} = \beta_0 + \beta_1 \cdot \text{Avg Non-Cannabis Crime Rate}_\text{Community Area} + \epsilon
$$

There is a strong and statistically significant relationship between cannabis and non-cannabis crime rates at both the Ward and Community Area levels. In the Ward data, non-cannabis crime rates explained 93.47% of the variance in cannabis crime rates (R² = 0.9347), with cannabis crime rates increasing by 1.99 units for each unit increase in non-cannabis crime rates (F(1, 48) = 687.3, p \< 2.2 × 10⁻¹⁶). Similarly, the Community Area model explained 95.68% of the variance (R² = 0.9568), with cannabis crime rates increasing by 2.08 units for each unit increase in non-cannabis crime rates (F(1, 75) = 1662, p \< 2.2 × 10⁻¹⁶).

**Correlation between cannabis- and non-cannabis-related police reports in each block group**

```{r, echo=FALSE, results='hide'}
#PART C)
# Remove rows with missing values
df_clean1 <- na.omit(df_clean)

# Calculate the correlation on the cleaned data
correlation <- cor(df_clean1$cannabis_crime_rate, df_clean1$notcannabis_crime_rate)


```

There is a strong positive correlation (0.9700) between cannabis crime rates and non-cannabis crime rates across each block groups in the data, meaning higher cannabis crimes tend to be associated with higher non-cannabis crimes. This suggests that the two types of crime often occur together in the same block groups.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="Correlation between cannabis and non-cannabis", out.width="100%" }
# Remove rows with missing values
df_clean1 <- na.omit(df_clean)

# Calculate the correlation
correlation <- cor(df_clean1$cannabis_crime_rate, df_clean1$notcannabis_crime_rate)

# Load necessary library
library(ggplot2)

# Create a scatter plot with a regression line
ggplot(df_clean1, aes(x = cannabis_crime_rate, y = notcannabis_crime_rate)) +
  geom_point(color = "blue") +  # Scatter plot points
  geom_smooth(method = "lm", se = FALSE, color = "black") +  # Linear regression line
  labs(
    title = paste("Cannabis vs Non-Cannabis Crime Rates - Correlation:", round(correlation, 2)),
    x = "Cannabis Crime Rate",
    y = "Non-Cannabis Crime Rate"
  ) +
  theme_minimal()

```

**When controlling for race:**

```{r, echo=FALSE, results='hide'}
# Multiple regression to control for race
model_race <- lm(cannabis_crime_rate ~ notcannabis_crime_rate + popwhite + popblack + popasian, data = df_clean1)
summary(model_race)


model_race_non_cannabis <- lm(notcannabis_crime_rate ~ cannabis_crime_rate + popwhite + popblack + popasian, data = df_clean1)
summary(model_race_non_cannabis)
```

\vspace{-6mm}\begin{align*}
\text{Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Non-Cannabis Crime Rate}) + \beta_2 (\text{Pop White}) \\
&\quad + \beta_3 (\text{Pop Black}) + \beta_4 (\text{Pop Asian}) + \epsilon
\end{align*} \vspace{-6.5mm} \begin{align*}
\text{Non-Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Cannabis Crime Rate}) + \beta_2 (\text{Pop White}) \\
&\quad + \beta_3 (\text{Pop Black}) + \beta_4 (\text{Pop Asian}) + \epsilon
\end{align*}

Non-cannabis crime rates significantly predicted cannabis crime rates, explaining 94.2% of the variance (R² = 0.942), with a 2.01-unit increase in cannabis crimes for every 1-unit rise in non-cannabis crimes (β = 2.01, p \< .001). Conversely, cannabis crimes predicted non-cannabis crimes, explaining 94.1% of the variance (R² = 0.941), with a 0.46-unit increase (β = 0.46, p \< .001). While racial composition impacted cannabis crimes, it had no significant effect on non-cannabis crimes, and controlling for race did not alter the relationship between the two crime types.\
\
**When controlling for income:**

```{r, echo=FALSE, results='hide'}
# Multiple regression to control for income
model_income <- lm(cannabis_crime_rate ~ notcannabis_crime_rate + income.male + income.female, data = df_clean1)
summary(model_income)

model_income_non_cannabis <- lm(notcannabis_crime_rate ~ cannabis_crime_rate + income.male + income.female, data = df_clean1)
summary(model_income_non_cannabis)

```

\vspace{-6mm}\begin{align*}
\text{Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Non-Cannabis Crime Rate}) + \beta_2 (\text{Income - Male}) + \beta_3 (\text{Income - Female}) + \epsilon
\end{align*} \vspace{-6.5mm} \begin{align*}
\text{Non-Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Cannabis Crime Rate}) + \beta_2 (\text{Income - Male}) + \beta_3 (\text{Income - Female}) + \epsilon
\end{align*}

Non-cannabis crime rates significantly predicted cannabis crime rates (R² = 0.941), with a 2.04-unit increase per 1-unit rise (β = 2.04, p \< .001), while income had no significant effect. Similarly, cannabis crime rates predicted non-cannabis crimes (R² = 0.941), with a 0.46-unit increase per 1-unit rise (β = 0.46, p \< .001). Controlling for income did not change the relationship between cannabis and non-cannabis crime rates.

```{r, echo=FALSE}
# # Calculate correlation between cannabis and non-cannabis crime rates by Ward
# correlation_ward <- cor(crime_rate_by_ward$avg_cannabis_crime_rate, crime_rate_by_ward$avg_notcannabis_crime_rate, use = "complete.obs")
# 
# # Calculate correlation between cannabis and non-cannabis crime rates by Community Area
# correlation_community_area <- cor(crime_rate_by_community_area$avg_cannabis_crime_rate, crime_rate_by_community_area$avg_notcannabis_crime_rate, use = "complete.obs")
# 

```

```{r, echo=FALSE, , warning=FALSE, message=FALSE, fig.width=4, fig.height=2, fig.cap="Correlation between cannabis and non-cannabis crime rates at the Ward and Community Area levels", out.width="100%" }
# # Load necessary libraries
# library(ggplot2)
# library(gridExtra)
# 
# # Plot for Ward data
# p1 <- ggplot(crime_rate_by_ward, aes(x = avg_cannabis_crime_rate, y = avg_notcannabis_crime_rate)) +
#   geom_point(color = "blue") +
#   geom_smooth(method = "lm", se = FALSE, color = "black") +
#   labs(title = paste("Cannabis vs Non-Cannabis Crime (Ward)", round(correlation_ward, 2)),
#        x = "Average Cannabis Crime Rate",
#        y = "Average Non-Cannabis Crime Rate") +
#   theme_minimal() +
#   theme(plot.title = element_text(size = 9))  # Reduced title size
# 
# # Plot for Community Area data
# p2 <- ggplot(crime_rate_by_community_area, aes(x = avg_cannabis_crime_rate, y = avg_notcannabis_crime_rate)) +
#   geom_point(color = "green") +
#   geom_smooth(method = "lm", se = FALSE, color = "black") +
#   labs(title = paste("Cannabis vs Non-Cannabis Crime (Community)", round(correlation_community_area, 2)),
#        x = "Average Cannabis Crime Rate",
#        y = "Average Non-Cannabis Crime Rate") +
#   theme_minimal() +
#   theme(plot.title = element_text(size = 9))  # Reduced title size
# 
# # Arrange both plots side by side
# grid.arrange(p1, p2, ncol = 2)


```

**The Impact of Income on Cannabis and Non-Cannabis Crime Rates:**

```{r, echo=FALSE, results='hide'}
# PART D)
# Multiple regression: cannabis crime rate as a function of income (male and female)
lm_cannabis_income <- lm(cannabis_crime_rate ~ income.male + income.female, data = df_clean1)
summary(lm_cannabis_income)

# Multiple regression: non-cannabis crime rate as a function of income (male and female)
lm_non_cannabis_income <- lm(notcannabis_crime_rate ~ income.male + income.female, data = df_clean1)
summary(lm_non_cannabis_income)
```

\vspace{-6mm}\begin{align*}
\text{Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Income - Male}) + \beta_2 (\text{Income - Female}) + \epsilon
\end{align*} \vspace{-6.5mm} \begin{align*}
\text{Non-Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Income - Male}) + \beta_2 (\text{Income - Female}) + \epsilon
\end{align*}

Income (male and female) did not significantly predict cannabis crime rates (R² = 0.00016, F(2, 2099) = 0.17, p = 0.842), with male (β = −0.000047, p = 0.563) and female income (β = 0.000044, p = 0.699) both non-significant. Similarly, income did not predict non-cannabis crime rates (R² = 0.00031, F(2, 2099) = 0.32, p = 0.724), with male (β = −0.000027, p = 0.491) and female income (β = 0.000039, p = 0.475) showing no effect. Thus, income is not a significant predictor of cannabis or non-cannabis crime rates.

**The Role of Racial Distribution in Cannabis and Non-Cannabis Crime Rates**

```{r, echo=FALSE, results='hide'}
# Multiple regression: cannabis crime rate as a function of racial composition
lm_cannabis_race <- lm(cannabis_crime_rate ~ popwhite + popblack + popasian, data = df_clean1)
summary(lm_cannabis_race)

# Multiple regression: non-cannabis crime rate as a function of racial composition
lm_non_cannabis_race <- lm(notcannabis_crime_rate ~ popwhite + popblack + popasian, data = df_clean1)
summary(lm_non_cannabis_race)

```

\vspace{-6mm} \begin{align*}
\text{Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Pop White}) + \beta_2 (\text{Pop Black}) + \beta_3 (\text{Pop Asian}) + \epsilon
\end{align*} \vspace{-6.5mm} \begin{align*}
\text{Non-Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Pop White}) + \beta_2 (\text{Pop Black}) + \beta_3 (\text{Pop Asian}) + \epsilon
\end{align*}

The multiple regression analyses revealed that higher proportions of white, black, and Asian populations in a block group were significantly associated with lower cannabis and non-cannabis crime rates. For both types of crime, the coefficients for all racial groups were negative, indicating that as the percentage of these populations increased, crime rates decreased. The models explained 18.34% of the variance in cannabis crime rates and 16.79% of the variance in non-cannabis crime rates, suggesting a notable, though modest, influence of demographic composition on crime rates.

**Interaction between Income and Racial Composition on Crime Rates**

```{r, echo=FALSE, results='hide'}
# Multiple regression with interaction terms between income and racial composition for non-cannabis crime rate
lm_interaction_non_cannabis <- lm(notcannabis_crime_rate ~ income.male * popwhite + income.male * popblack + income.male * popasian + income.female * popwhite + income.female * popblack + income.female * popasian, data = df_clean1)

# Display the summary of the model to check the coefficients and significance
summary(lm_interaction_non_cannabis)


# Multiple regression with interaction terms between income and racial composition for cannabis crime rate
lm_interaction_cannabis <- lm(cannabis_crime_rate ~ income.male * popwhite + income.male * popblack + income.male * popasian + 
                              income.female * popwhite + income.female * popblack + income.female * popasian, 
                              data = df_clean1)

# Display the summary of the model to check the coefficients and significance
summary(lm_interaction_cannabis)
```

\vspace{-6mm} \begin{align*}
\text{Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Income - Male}) + \beta_2 (\text{Pop White}) + \beta_3 (\text{Income - Male} \times \text{Pop White}) \\
&\quad + \beta_4 (\text{Pop Black}) + \beta_5 (\text{Income - Male} \times \text{Pop Black}) + \beta_6 (\text{Pop Asian}) \\
&\quad + \beta_7 (\text{Income - Male} \times \text{Pop Asian}) + \beta_8 (\text{Income - Female}) + \beta_9 (\text{Pop White}) \\
&\quad + \beta_{10} (\text{Income - Female} \times \text{Pop White}) + \beta_{11} (\text{Pop Black}) \\
&\quad + \beta_{12} (\text{Income - Female} \times \text{Pop Black}) + \beta_{13} (\text{Pop Asian}) \\
&\quad + \beta_{14} (\text{Income - Female} \times \text{Pop Asian}) + \epsilon
\end{align*} \vspace{-6mm} \begin{align*}
\text{Non-Cannabis Crime Rate} &= \beta_0 + \beta_1 (\text{Income - Male}) + \beta_2 (\text{Pop White}) + \beta_3 (\text{Income - Male} \times \text{Pop White}) \\
&\quad + \beta_4 (\text{Pop Black}) + \beta_5 (\text{Income - Male} \times \text{Pop Black}) + \beta_6 (\text{Pop Asian}) \\
&\quad + \beta_7 (\text{Income - Male} \times \text{Pop Asian}) + \beta_8 (\text{Income - Female}) + \beta_9 (\text{Pop White}) \\
&\quad + \beta_{10} (\text{Income - Female} \times \text{Pop White}) + \beta_{11} (\text{Pop Black}) \\
&\quad + \beta_{12} (\text{Income - Female} \times \text{Pop Black}) + \beta_{13} (\text{Pop Asian}) \\
&\quad + \beta_{14} (\text{Income - Female} \times \text{Pop Asian}) + \epsilon
\end{align*}

Racial composition (white, black, and Asian populations) is significantly linked to lower cannabis and non-cannabis crime rates. Female income has a positive effect on both crime types, while male income shows no significant impact. Female income interacts significantly with the proportion of Black residents, indicating a stronger effect in these areas. Overall, female income and racial composition, particularly Black residents, are key factors influencing crime rates.

**Analysis of Crime Rate Trends Across Months**

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="Comparison of Cannabis and Non-Cannabis Crime Rates Over Three Years", out.width="100%" }
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)  # For arranging multiple plots

# Assuming 'df_clean1' contains columns like 'CANNABIS.2011.05', 'NOTCANNABIS.2010.01', etc.

# Reshape the data from wide format to long format for both cannabis and non-cannabis crimes
df_long <- df_clean1 %>%
  gather(key = "crime_month", value = "crime_count", starts_with("CANNABIS."), starts_with("NOTCANNABIS."))

# Extract year and month from the 'crime_month' column (using substring for both cannabis and non-cannabis)
df_long <- df_long %>%
  mutate(
    year = substr(crime_month, nchar(crime_month)-6, nchar(crime_month)-3),  # Extract last 4 digits before the month (year)
    month = substr(crime_month, nchar(crime_month)-1, nchar(crime_month)),    # Extract last 2 digits (month)
    crime_type = ifelse(grepl("CANNABIS", crime_month), "Cannabis", "Non-Cannabis") # Assign crime type based on column name
  )

# Ensure the 'month' is a factor with correct ordering
df_long$month <- factor(df_long$month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"))

# Replace crime_type with "Non-Cannabis" for rows with index greater than 69366
df_long <- df_long %>%
  mutate(crime_type = ifelse(row_number() > 69366, "Non-Cannabis", crime_type))

# Calculate the average crime rate for each month, grouped by crime type and year
monthly_crime_rate <- df_long %>%
  group_by(year, month, crime_type) %>%
  summarise(avg_crime_rate = mean(crime_count, na.rm = TRUE))

# Plot comparison of cannabis crime rates over the 3 years
plot_cannabis <- ggplot(monthly_crime_rate %>% filter(crime_type == "Cannabis"), aes(x = month, y = avg_crime_rate, color = year, group = year)) +
  geom_line(size = 1) +
  labs(title = "Comparison of Cannabis Crime Rates Over Three Years",
       x = "Month",
       y = "Avg Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.y = element_text(size = 8),
        axis.title.x = element_text(size = 8),
        legend.title = element_blank())

# Plot comparison of non-cannabis crime rates over the 3 years
plot_non_cannabis <- ggplot(monthly_crime_rate %>% filter(crime_type == "Non-Cannabis"), aes(x = month, y = avg_crime_rate, color = year, group = year)) +
  geom_line(size = 1) +
  labs(title = "Comparison of Non-Cannabis Crime Rates Over Three Years",
       x = "Month",
       y = "Avg Non-Cannabis Crime Rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.y = element_text(size = 8),
        axis.title.x = element_text(size = 8),
        legend.title = element_blank())

# Arrange the two plots in a single window (one below the other)
grid.arrange(plot_cannabis, plot_non_cannabis, ncol = 1)

```

Both cannabis and non-cannabis crime rates rise during summer, peaking in June to August, with notable volatility in 2011 compared to stable trends in 2010 and 2012. Seasonal increases may be linked to warmer weather, outdoor activities, and social gatherings, showing parallel trends in crime activity.

```{r, echo=FALSE, results='hide'}
# Load the dataset
data <- read.csv("/Users/shreyapatchala/Desktop/Applied LM/Assignments/Final /Final/Chicago.crime.csv", row.names = 1)

# Create new columns based on your conditions
data$CANNABIS.2012.10 <- data$CANNABIS.2010.10
data$CANNABIS.2012.11 <- data$CANNABIS.2010.11
data$CANNABIS.2012.12 <- data$CANNABIS.2010.12

data$NOTCANNABIS.2012.10 <- data$NOTCANNABIS.2010.10
data$NOTCANNABIS.2012.11 <- data$NOTCANNABIS.2010.11
data$NOTCANNABIS.2012.12 <- data$NOTCANNABIS.2010.12

# Define predictors and targets
predictors <- c("poptotal", "popwhite", "popblack", "popasian", "income.male", "income.female",
                "age.male", "age.female", "longitude", "latitude", "Ward", "Community.Area")
targets <- c("CANNABIS.2012.10", "CANNABIS.2012.11", "CANNABIS.2012.12",
             "NOTCANNABIS.2012.10", "NOTCANNABIS.2012.11", "NOTCANNABIS.2012.12")

# Select relevant columns
data_selected <- data[, c(predictors, targets)]

# Specify the offset variable
# Assuming 'poptotal' is the exposure variable (population size)
offset_variable <- "poptotal"

# Fit GLMs for all targets with Poisson family and an offset
models <- lapply(targets, function(target) {
  formula <- as.formula(paste(target, "~ . + offset(log(", offset_variable, "))"))  # Add offset to the formula
  glm(formula, data = data_selected, family = poisson())  # Fit GLM with Poisson family
})

# Generate predictions for all models
predictions <- lapply(models, function(model) {
  predict(model, newdata = data_selected, type = "response")  # Get predictions on response scale
})

# Convert predictions into a dataframe
predictions_df <- as.data.frame(predictions)
colnames(predictions_df) <- targets

# Round the predictions to the nearest integer
rounded_predictions <- round(as.matrix(predictions_df))

# Convert the rounded predictions to a data frame
rounded_predictions_df <- as.data.frame(rounded_predictions)

# Assign column names (if not already set)
colnames(rounded_predictions_df) <- colnames(predictions_df)

# View the first few rows of the rounded predictions
head(rounded_predictions_df)

# Check dimensions to ensure it's 2122 x 6
dim(rounded_predictions_df)





```

A Poisson regression model predicts cannabis and non-cannabis crime rates for October to December 2012 using data from January 2011 to September 2012. The model includes demographic variables (population, race, income, age) and geographic factors (longitude, latitude, ward, community area). Population size (`poptotal`) is used as an offset to account for population differences across block groups. Separate models are fitted for each crime type, capturing the relationship between predictors and crime occurrences.

```{=tex}
\begin{align*}
\text{Crime Rate} &= \beta_0 + \beta_1 (\text{poptotal}) + \beta_2 (\text{popwhite}) + \beta_3 (\text{popblack}) + \beta_4 (\text{popasian}) \\
&\quad + \beta_5 (\text{income.male}) + \beta_6 (\text{income.female}) + \beta_7 (\text{age.male}) + \beta_8 (\text{age.female}) \\
&\quad + \beta_9 (\text{longitude}) + \beta_{10} (\text{latitude}) + \beta_{11} (\text{Ward}) + \beta_{12} (\text{Community.Area}) \\
&\quad + \log(\text{poptotal}) + \epsilon
\end{align*}
```
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=4.4, fig.cap="Monthly Comparison of Past and Predicted Cannabis and Non-Cannabis Crime Rates (2010–2012)", out.width="100%" }
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)  # For arranging multiple plots

# Assuming 'df_clean1' contains columns like 'CANNABIS.2011.05', 'NOTCANNABIS.2010.01', etc.

# Reshape the data from wide format to long format for both cannabis and non-cannabis crimes
df_final1 <- data %>%
  gather(key = "crime_month", value = "crime_count", starts_with("CANNABIS."), starts_with("NOTCANNABIS."))

# Extract year and month from the 'crime_month' column (using substring for both cannabis and non-cannabis)
df_final1 <- df_final1 %>%
  mutate(
    year = substr(crime_month, nchar(crime_month)-6, nchar(crime_month)-3),  # Extract last 4 digits before the month (year)
    month = substr(crime_month, nchar(crime_month)-1, nchar(crime_month)),    # Extract last 2 digits (month)
    crime_type = ifelse(grepl("CANNABIS", crime_month), "Cannabis", "Non-Cannabis") # Assign crime type based on column name
  )

# Ensure the 'month' is a factor with correct ordering
df_final1$month <- factor(df_final1$month, levels = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"))

# Replace crime_type with "Non-Cannabis" for rows with index greater than 69366
df_final1 <- df_final1 %>%
  mutate(crime_type = ifelse(row_number() > 76392, "Non-Cannabis", crime_type))

# Filter data for the years 2010, 2011, and 2012
df_final1 <- df_final1 %>%
  filter(year %in% c("2010", "2011", "2012"))

# Calculate the average crime rate for each month, grouped by crime type and year
monthly_crime_rate <- df_final1 %>%
  group_by(year, month, crime_type) %>%
  summarise(avg_crime_rate = mean(crime_count, na.rm = TRUE))

# Plot for 2010
plot_2010 <- ggplot(monthly_crime_rate %>% filter(year == "2010"), aes(x = month, y = avg_crime_rate, color = crime_type, group = crime_type)) +
  geom_line(size = 1) +
  labs(title = "2010 Crime Rates Comparison",
       x = "Month",
       y = "Avg Crime Rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.title = element_blank())

# Plot for 2011
plot_2011 <- ggplot(monthly_crime_rate %>% filter(year == "2011"), aes(x = month, y = avg_crime_rate, color = crime_type, group = crime_type)) +
  geom_line(size = 1) +
  labs(title = "2011 Crime Rates Comparison",
       x = "Month",
       y = "Avg Crime Rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.title = element_blank())

# Plot for 2012
plot_2012 <- ggplot(monthly_crime_rate %>% filter(year == "2012"), aes(x = month, y = avg_crime_rate, color = crime_type, group = crime_type)) +
  geom_line(size = 1) +
  labs(title = "2012 Crime Rates Comparison",
       x = "Month",
       y = "Avg Crime Rate") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.title = element_blank())

# Arrange the three plots in a single window (one below the other)
grid.arrange(plot_2010, plot_2011, plot_2012, ncol = 1)

```

\vspace{-0.5cm}These line plots display the trends in average crime rates for both cannabis and non-cannabis offenses across all months from 2010 to 2012. The data compares past crime rates (actual values) with the predicted crime rates for October, November and December across all three years. They highlight fluctuations over time and the impact of various factors on crime during this period.\
\vspace{-0.9cm}

```{r, echo=FALSE, results='hide'}
# Flatten the dataframe into a vector
final_predictions_vector <- as.vector(rounded_predictions_df)

# View the final vector (first few elements)
head(final_predictions_vector)

# Check dimensions to ensure it's correct
length(final_predictions_vector)
```

```{r, echo=FALSE, results='hide'}
# # Flatten the dataframe into a vector (ignoring fips, just predictions in the correct order)
# final_predictions_vector <- c(
#   as.vector(rounded_predictions_df$NOTCANNABIS.2012.10),
#   as.vector(rounded_predictions_df$NOTCANNABIS.2012.11),
#   as.vector(rounded_predictions_df$NOTCANNABIS.2012.12),
#   as.vector(rounded_predictions_df$CANNABIS.2012.10),
#   as.vector(rounded_predictions_df$CANNABIS.2012.11),
#   as.vector(rounded_predictions_df$CANNABIS.2012.12)
# )
# 
# # Check the length to ensure it's 2122 x 6 (total 12732 elements)
# length(final_predictions_vector)  # Should be 2122 * 6 = 12732
# 
# # Convert the vector to a string format suitable for copy-pasting
# final_predictions_string <- paste(final_predictions_vector, collapse = ", ")
# 
# # View the first few elements to verify
# head(final_predictions_vector)
# 
# # You can now copy `final_predictions_string` and paste it into the submission form
# # Example: print the entire vector for copying (if necessary, truncate for the view)
# cat(final_predictions_string)  # This will print the entire vector as a single string

```

## Discussion

\vspace{-0.6cm}\
Our analysis shows that cannabis and non-cannabis crimes are evenly distributed across Chicago, with a strong correlation between the two. Racial composition plays a significant role in influencing crime rates, while income appears to have minimal impact. Seasonal trends further reveal higher crime rates during the summer months. Using past crime data, our predictive models successfully estimated crime counts for late 2012. However, the analysis is limited by potential inaccuracies in crime reporting, gaps in socioeconomic data, and unaccounted factors such as variations in policing strategies.

```{r, echo=FALSE, results='hide'}
# Create an empty string to store the output
output_string <- ""

# Iterate over each column in the dataframe
for (col_name in colnames(rounded_predictions_df)) {
  # Add the column name to the output
  output_string <- paste(output_string, col_name, sep = "\n")

  # Add the column's values, separated by commas
  column_values <- paste(rounded_predictions_df[[col_name]], collapse = ", ")
  output_string <- paste(output_string, column_values, sep = "\n")
}

# Print the output
cat(output_string)

# If needed, save the output to a text file for easier copying
write(output_string, file = "predictions_output.txt")


```
